---
title: "cyclistic R markdown"
output:
  html_document: default
date: "2022-11-30"
---

## install and loading packages
```{r}
library(magrittr)
library(dplyr)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(readr)
library(data.table)
```

## improting .csv files
```{r}
trips_2021_01 <- read.csv("./data/202101_divvy_tripdata.csv")
trips_2021_02 <- read.csv("./data/202102_divvy_tripdata.csv")
trips_2021_03 <- read.csv("./data/202103_divvy_tripdata.csv")
trips_2021_04 <- read.csv("./data/202104_divvy_tripdata.csv")
trips_2021_05 <- read.csv("./data/202105_divvy_tripdata.csv")
trips_2021_06 <- read.csv("./data/202106_divvy_tripdata.csv")
trips_2021_07 <- read.csv("./data/202107_divvy_tripdata.csv")
trips_2021_08 <- read.csv("./data/202108_divvy_tripdata.csv")
trips_2021_09 <- read.csv("./data/202109_divvy_tripdata.csv")
trips_2021_10 <- read.csv("./data/202110_divvy_tripdata.csv")
trips_2021_11 <- read.csv("./data/202111_divvy_tripdata.csv")
trips_2021_12 <- read.csv("./data/202112_divvy_tripdata.csv")
```

## Compare column names each of the files
```{r}
colnames(trips_2021_01)
colnames(trips_2021_02)
colnames(trips_2021_03)
colnames(trips_2021_04)
colnames(trips_2021_05)
colnames(trips_2021_06)
colnames(trips_2021_07)
colnames(trips_2021_08)
colnames(trips_2021_09)
colnames(trips_2021_10)
colnames(trips_2021_11)
colnames(trips_2021_12)
```
## Inspect the dataframes and look for incongruencies
```{r}
str(trips_2021_01)
str(trips_2021_02)
str(trips_2021_03)
str(trips_2021_04)
str(trips_2021_05)
str(trips_2021_06)
str(trips_2021_07)
str(trips_2021_08)
str(trips_2021_09)
str(trips_2021_10)
str(trips_2021_11)
str(trips_2021_12)
```

## Stack individual data frames into one big data frame
```{r}
all_trips <- bind_rows(trips_2021_01, trips_2021_02, trips_2021_03, trips_2021_04,
                       trips_2021_05, trips_2021_06, trips_2021_07,
                       trips_2021_08, trips_2021_09, trips_2021_10, trips_2021_11,
                       trips_2021_12)
```
## remove data from environment
```{r}
remove(trips_2021_01, trips_2021_02, trips_2021_03, trips_2021_04, trips_2021_05,
       trips_2021_06, trips_2021_07, trips_2021_08, trips_2021_09, trips_2021_10,
       trips_2021_11, trips_2021_12)
```
## Inspect the new table that has been created
```{r}
colnames(all_trips)
nrow(all_trips)
dim(all_trips)
head(all_trips)
str(all_trips)
summary(all_trips)
```

## Remove lat, long, and gender fields
```{r}
all_trips <- all_trips%>% 
  select(-c(start_lat, start_lng, end_lat, end_lng))
```

## Remove null and blank rows
```{r}
all_trips <- all_trips[!(is.na(all_trips$start_station_name) | all_trips$start_station_name=="" 
                          | is.na(all_trips$start_station_id) | all_trips$start_station_id==""
                          |is.na(all_trips$end_station_name) | all_trips$end_station_name==""
                          |is.na(all_trips$end_station_id) | all_trips$end_station_id==""),]
```
## change format for started_at and ended_at chr to date format
```{r}
all_trips$started_at <-lubridate::ymd_hms(all_trips$started_at)
all_trips$ended_at <-lubridate::ymd_hms(all_trips$ended_at)
```

## Add columns that list the date, month and day of each ride
```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%b")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```
## Add a "ride_length" calculation to all_trips (in seconds)
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```
## Convert "ride_length" from Factor to numeric so we can run calculations on the data
```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```
## Remove "bad" data
```{r}
all_trips <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```
## Descriptive analysis on ride_length (all figures in seconds)
```{r}
mean(all_trips$ride_length)
median(all_trips$ride_length)
max(all_trips$ride_length)
min(all_trips$ride_length)
```

## You can condense the four lines above to one line using summary() on the specific attribute
```{r}
summary(all_trips$ride_length)
```

## Compare members and casual users
```{r}
aggregate(all_trips$ride_length~all_trips$member_casual, FUN = mean)
aggregate(all_trips$ride_length~all_trips$member_casual, FUN = median)
aggregate(all_trips$ride_length~all_trips$member_casual, FUN = max)
aggregate(all_trips$ride_length~all_trips$member_casual, FUN = min)
```

## See the average ride time by each day for members vs casual users
```{r}
aggregate(all_trips$ride_length~all_trips$member_casual + all_trips$day_of_week, FUN = mean)
```

## the days of the week are out of order. Let's fix that
```{r}
all_trips$day_of_week <- ordered(all_trips$day_of_week, levels=c("Sunday", 
                                                                       "Monday", 
                                                                       "Tuesday", 
                                                                       "Wednesday", 
                                                                       "Thursday", 
                                                                       "Friday", 
                                                                       "Saturday"))
```
## let's see the weekday show in order
```{r}
aggregate(all_trips$ride_length~all_trips$member_casual + all_trips$day_of_week, FUN = mean)
```
## analyze ridership data by type and weekday
```{r}
all_trips %>%  
  mutate(weekday = lubridate::wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)
```

## visualize the number of rides by rider type
```{r}
all_trips %>%  
  mutate(weekday = lubridate::wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x=weekday, y = number_of_rides, fill = member_casual))+
     geom_col(position = "dodge")
```

## visualization for average duration
```{r}
all_trips %>% 
  mutate(weekday = lubridate::wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual))+
  geom_col(position = "dodge")
```
## create new table of average ride length with column name member_casual, day of week, ride length
```{r}
counts <- aggregate(all_trips$ride_length ~ all_trips$member_casual + all_trips$day_of_week, FUN = mean)
```
## Create a csv file that we will visualize in Excel, Tableau, or my presentation software
```{r}
write.csv(counts, file = 'F:/Rstudio Project/Cyclistic/cyclistic/avg_ride_length.csv')
write.csv(all_trips, file = 'F:/Rstudio Project/Cyclistic/cyclistic/all_trips.csv')
```

